---
title: "Simulating Poisson processes"
subtitle: "University of Toronto, ACT 350"
author: "Prof. S. Pesenti"
date: "19/10/2019"

header-includes:
- \usepackage{lastpage}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \setlength{\headheight}{23pt}
- \fancyhead[L]{ACT 350}
- \fancyhead[R]{University of Toronto\\ Prof. S. Pesenti}
- \fancyfoot[C]{}
- \fancyfoot[R]{\thepage \, of \pageref{LastPage}}
output:
  pdf_document: default
  html_document: default
---
```{r packages, include=FALSE}
# package for calculating matrix exponentials
```


# How to simulate a Poisson process


Recall that a Poisson process with rate $\lambda >0$ is a counting process $\{ N(t), ~t \geq 0\}$ that has the following properties


1. *starts at 0*: $N(0) = 0$ with probability 1    
2. *is increasing*: $N(s)  \leq N(t)$ for all $s \leq t$    
3. *has Poisson distributed increments*: $P(N(s + t) - N(s) = n) \sim$ Pois($\lambda t$) for all $s,t \geq 0$.     


However, this definition does not help if we want to simulate a Poisson process. Thus, to simulate a Poisson process we use its alternative definition. That is, for independent random variables $X_1, X_2, \ldots$ that are Exponentially distributed with rate $\lambda>0$, a Poisson process with rate $\lambda>0$ can be written as
\begin{equation*}
N(t) = \max\Big\{ n \in \mathbb{N} ~\Big|~ \sum_{i = 1}^n X_i \leq t \Big\}, \quad \text{for all } t \geq 0.
\end{equation*}


Thus, for simulating a Poisson process we first need to simulate exponential random variables, add them up and check whether they are smaller or equal to $t$. However, we will run into problems when we want to store the value of $N(t)$ for every $0 \leq t < \infty$. The way out is to implement $N(t)$ as a function of $t$ in the following way. First generate the jump points of the Poisson process. Second, we know that a Poisson process can only jump one unit at a time. Thus, we can define the Poisson process as a step function that jumps exactly one unit at every jump point. 


## 1. Generating jump times
This is an algorithm for generating the jump times of a Poisson process    


1. Set $\tau_0 = 0$.     
2. For $k \geq 1$ do     
    a) generate $X \sim$Exp($\lambda$)      
    b) set $\tau_k = \tau_{k - 1} + X$    

The $\tau_k, ~ k = 0, 1, \ldots$ are the time points, where the Poisson process jumps. 


**Numerical implementation: R code**
```{r}
# set the seed for generating random variables (for reproducability) 
# set.seed(2019)

# set the rate of the Poisson process/Exponential distribution
lambda <- 2
# number of jumps 
n <- 50
# initialise the jump times
jump_times <- rep(0, lenthought = n)

# note that the first value of the jump_times is aleady 0
for(i in 2:n){
  jump_times[i] <- jump_times[i - 1] + rexp(1, lambda)
}
jump_times
```


A more efficient implementation, avoiding the for loop, is 
```{r}
jump_times_2 = c(0, cumsum(rexp(n-1, lambda)))
# note that the two generated sample_paths are different, 
# since the Exponential random variables are different.
jump_times_2
```

## 2. Defining the Poisson process as a function

Next, we can define the Poisson process as a step function that jumps exactly one unit at every value in `jump_times`. Note that the implemented Poisson process is now a *function* of the time $t$.   
```{r}
# define the Poisson process as a function
poisson_process <- stepfun(jump_times, seq(0, n, by = 1))
# the first argument of `stepfun` is where the step function jumps
# the second argument is the hight of the step function

# Let us check some values.
# What is the value of N(3)? (Note that we use jump_times)
poisson_process(3)
poisson_process(10)
poisson_process(15)

```


Note you can evaluate the Poisson process for any time point $t$, however, keep in mind that we only simulated 50 jumps of the Poisson process. Thus, evaluating `N(t)` for $t$ large does not make sense. 

## 3. Plotting a sample path

To plot a sample path we have to plot the implemented step function `poisson_process`. 
```{r, simulated sample path}
plot(poisson_process, xlab = "time", ylab = "Value of Poisson process", main = NULL,
     verticals = FALSE, do.points = FALSE, xlim = c(0,jump_times[n]))

```

**Question:** Rerun the code, what do you observe?     
**Question:** Rerun the code with a different $\lambda$ or change the number of jumps. How does the sample path change?      
**Question:** What happens if you replace the Exponential random variables in the definition of the Poisson process with other i.i.d. random variables? For example, Gamma, Normal or LogNormal?    
Rerun the code snippets changing `rexp` to `rgamma`, `rnorm` or `rlnorm`.    

*Note* that for random variables that are not Exponentially distributed you still get a counting process. These processes are called *Renewal processes*. 









