---
title: "Simulating Poisson processes"
subtitle: "University of Toronto, ACT 350"
author: "Prof. S. Pesenti"
date: "08/09/2020"
header-includes:
- \usepackage{lastpage}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \setlength{\headheight}{23pt}
- \fancyhead[L]{ACT 350}
- \fancyhead[R]{University of Toronto\\ Prof. S. Pesenti}
- \fancyfoot[C]{}
- \fancyfoot[R]{\thepage \, of \pageref{LastPage}}
output:
  pdf_document: default
  html_document: default
  md_document: default
---
```{r packages, include=FALSE}
# package for calculating matrix exponentials
```


# How to simulate a Poisson process


Recall that a Poisson process with rate $\lambda >0$ is a counting process $\{ N(t), ~t \geq 0\}$ that has the following properties


1. *starts at 0*: $N(0) = 0$ with probability 1    
2. *is increasing*: $N(s)  \leq N(t)$ for all $s \leq t$    
3. *has Poisson distributed increments*: $P(N(s + t) - N(s) = n) \sim$ Pois($\lambda t$) for all $s,t \geq 0$.     


However, this definition does not help if we want to simulate a Poisson process. Thus, to simulate a Poisson process we use its alternative definition. That is, for independent random variables $X_1, X_2, \ldots$ that are Exponentially distributed with rate $\lambda>0$, a Poisson process with rate $\lambda>0$ can be written as
\begin{equation*}
N(t) = \max\Big\{ n \in \mathbb{N} ~\Big|~ \sum_{i = 1}^n X_i \leq t \Big\}, \quad \text{for all } t \geq 0.
\end{equation*}


Thus, for simulating a Poisson process we first need to simulate exponential random variables, add them up and check whether they are smaller or equal to $t$. However, we will run into problems when we want to store the value of $N(t)$ for every $0 \leq t < \infty$. The way out is to implement $N(t)$ as a function of $t$ in the following way. First generate the jump points of the Poisson process. Second, we know that a Poisson process can only jump one unit at a time. Thus, we can define the Poisson process as a step function that jumps exactly one unit at every jump point. 


## 1. Generating jump times
This is an algorithm for generating the jump times of a Poisson process    


1. Set $\tau_0 = 0$.     
2. For $k \geq 1$ do     
    a) generate $X \sim$Exp($\lambda$)      
    b) set $\tau_k = \tau_{k - 1} + X$    

The $\tau_k, ~ k = 0, 1, \ldots$ are the time points, where the Poisson process jumps. 


**Numerical implementation: R code**
```{r}
# set the seed for generating random variables (for reproducability) 
# set.seed(2019)

# set the rate of the Poisson process/Exponential distribution
lambda <- 2
# number of jumps 
n <- 50
# initialise the jump times
jump_times <- rep(0, lenthought = n)

# note that the first value of the jump_times is aleady 0
for(i in 2:n){
  jump_times[i] <- jump_times[i - 1] + rexp(1, lambda)
}
jump_times
```


A more efficient implementation, avoiding the for loop, is 
```{r}
jump_times_2 = c(0, cumsum(rexp(n-1, lambda)))
# note that the two generated sample_paths are different, 
# since the Exponential random variables are different.
jump_times_2
```

## 2. Defining the Poisson process as a function

Next, we can define the Poisson process as a step function that jumps exactly one unit at every value in `jump_times`. Note that the implemented Poisson process is now a *function* of the time $t$.   
```{r}
# define the Poisson process as a function
poisson_process <- stepfun(jump_times[2 : 50], seq(0, n - 1, by = 1))
# the first argument of `stepfun` is where the step function jumps
# the second argument is the hight of the step function

# Let us check some values.
# The starting value is 0.
poisson_process(0)
# What is the value of N(3)? (Note that we use jump_times and not jump_times_2)
poisson_process(3)
poisson_process(10)
poisson_process(15)
```

The first jump time is `r jump_times[2]`. Let us check whether the Poisson 
process jumps at that time.

```{r}
# The first jump of the Poisson process is 
first_jump <- jump_times[2]
poisson_process(first_jump - 0.001)
poisson_process(first_jump + 0.001)
```

Note you can evaluate the Poisson process for any time point $t$, however, keep in mind that we only simulated 50 jumps of the Poisson process. Thus, evaluating `N(t)` for $t$ large does not make sense. 

## 3. Plotting a sample path

To plot a sample path we have to plot the implemented step function `poisson_process`. 
```{r, simulated sample path}
plot(poisson_process, xlab = "time", ylab = "Value of Poisson process", main = NULL,
     verticals = FALSE, do.points = FALSE, xlim = c(0,jump_times[n]))

```

**Question:** Rerun the code, what do you observe?     
**Question:** Rerun the code with a different $\lambda$ or change the number of jumps. How does the sample path change?      


# Renewal Theory

A renewal process is defined as follows: Consider a sequence $T_1, T_2, \ldots$ of independent identically distributed (i.i.d.) non-negative random variables. Define the stochastic process
$$ X_0 = 0\,, \quad X_n = T_1 + \dots + T_n\,, \quad  n = 1, 2, \dots\,,$$
and set 
$$N(t) = \max\Big\{ n \in \mathbb{N} ~\Big|~ \sum_{i = 1}^n X_i \leq t \Big\}, \quad \text{for all } t \geq 0. $$
Then the process $\{N(t)\, |\, t \ge 0\}$ is called a renewal process. 

If the $T_i\, ,i = 1, 2, \ldots$ are i.i.d. Exponential with parameter $\lambda>0$, then the renewal process $\{N(t)\,|\, t \ge 0\}$ is a Poisson process. 


# **Task:**    

**a)** Generate a code to simulate from a renewal process, where the interarrival times $T_i\, ,i = 1, 2, \ldots$ are 
i.i.d. LogNormal distributed with mean $2$ and standard deviation 0.5. *Hint: use the function `rlnorm`*.


**b)** Plot a sample path of the renewal process define in **a)**. 

**c)** What do you observe compared to the Poisson process? 

**d)** Repeat tasks **a)** to **b)** with interarrival times being Gamma distributed with the same mean and standard deviation as in **a)**. *Hint: use the function `rgamma`*.

**e)** Compare the Poisson process with the renewal process from **a)** and the renewal process from **d)** and give interpretations. 











